Discrete Quantum Theories and Computing

### 2018-03-09 dissertation.tex
dissertation.lyx "2.1.3 Explicit generalization of the Hopf fibration construction" Cleaned up...

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Hat{a}$ and $\Hat{b}$ on $\Sphere{2}$,
the shortest rotation carrying $\Hat{a}$ to $\Hat{b}$ is the SLERP
(spherical linear interpolation) \cite{Shoemake:1985:ARQ:325334.325242,wiki:Slerp}
\begin{equation}
S(\Hat{a},\Hat{b},t)=\Hat{a}\,\frac{\sin((1-t)\omega)}{\sin\omega}\,+\,\Hat{b}\,\frac{\sin(t\omega)}{\sin\omega}\ ,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\omega$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

@@

\subsection{Explicit $n$-qubit generalization of the Hopf fibration construction}

For one qubit, we could easily solve the problem of reducing the full
unit-norm space to its irreducible components $\Vec{X}=(X,Y,Z)$ characterizing
the Bloch sphere. We have just argued that essentially the same process
is possible for $n$-qubits: in the abstract argument, we simply identify
the family of coefficients $\{\alpha_{i}\}$ as being the same if
they differ only by an overall phase $\rme^{\rmi\theta}$. However,
in practice this is not a construction that is easy to realize in
a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $n$-qubit state space
$\CP{D-1}$; this construction will turn out to be useful for the
validation of our discrete results to follow below.

We begin by noting that a natural quantity characterizing an $n$-qubit
system is its \textit{density matrix\/}, $\rho=\left[\alpha_{i}\alpha_{j}^{*}\right]$,
or 
\begin{equation}
\rho=\left[\begin{array}{cccc}
|\alpha_{0}|^{2} & \alpha_{0}\alpha_{1}^{*} & \cdots & \alpha_{0}\alpha_{D-1}^{*}\\
\alpha_{1}\alpha_{0}^{*} & |\alpha_{1}|^{2} & \cdots & \alpha_{1}\alpha_{D-1}^{*}\\
\vdots & \vdots & \ddots & \vdots\\
\alpha_{D-1}\alpha_{0}^{*} & \cdots & \alpha_{D-1}\alpha_{D-2}^{*} & |\alpha_{D-1}|^{2}
\end{array}\right]\ .\label{nqDensity.eq}
\end{equation}

We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\theta}$ from the $n$-qubit states. If we take
a particular weighting of the elements of the density matrix $\rho$,
we can construct a \textit{unit vector\/} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Vec{X}=(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,\label{nqCoords.eq}
\end{equation}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
This construction gives an explicit embedding of the $(D-1)$-dimensional
complex, or $(2D-2)=(2^{n+1}-2)$-dimensional real, object in a real
space of dimension $D^{2}=2^{2n}$. However, this is somewhat subtle
because the vector is of unit length, so technically the embedding
space is a sphere of dimension $D^{2}-1=2^{2n}-1$ embedded in $\R^{D^{2}}$;
the one-qubit irreducible states could be represented in a 4D embedding,
but the magnitude of every coordinate would be one; furthermore, the
object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two\/} coordinate patches, one covering one pole of
$\Sphere{2}$ with coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & 1+0i\\
\alpha_{1} & = & x_{1}+iy_{1}
\end{array}
\end{eqnarray}
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates 
\begin{eqnarray}
\begin{array}{ccc}
\alpha_{0} & = & x_{0}+iy_{0}\\
\alpha_{1} & = & 1+0i\ .
\end{array}
\end{eqnarray}

We finally see that the irreducible $n$-qubit state space $\CP{D-1}$
is described by $D$ projectively equivalent coordinates, one of which
can always be scaled out to leave $(D-1)$ actual (complex) degrees
of freedom. We must choose, in turn, $D$ different local sets of
complex variables defined by taking the value $\alpha_{k}=1$, with
$k=0,\ldots,D-1$, and allowing the remaining $D-1$ complex (or $2D-2$
real) variables to run free. No single set of coordinates will work,
since the submanifold including $\alpha_{k}=0$ is undefined and another
coordinate system must be chosen to cover that coordinate patch. This
is a standard feature of the topology of non-trivial manifolds such
as $\CP{D-1}$ (see any textbook on geometry \cite{Berger}).

@@

This process of describing the discrete $n$-qubit irreducible states
can again be understood geometrically by following the discrete analog
of the Hopf fibration. First, we construct the discrete version of
the quadratic unit-length form that automatically annihilates the
distinction among states differing only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Vec{X}=(|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
{}From \ref{appH.sec}, we know that $p+1$ elements of this discrete
$\Sphere{2\times2^{n}-1}$ structure map to the \textit{same point\/}
in $\Vec{X}$. Each set of $(p+1)$ redundant points is, geometrically
speaking, the \textit{discrete Hopf fibration circle\/} living above
each \textit{irreducible\/} point of the $n$-qubit state description.
These $p+1$ points are interpretable as the $p$ finite points plus
the single point at infinity of the projective discrete line (see,
e.g., \cite{Arnold}).

@@

Finally, we repeat the last step of the $n$-qubit continuous Hopf
fibration process for discrete $n$-qubit states, eliminating the
discrete set of $p+1$ equivalent points that map to the same point
$\Vec{X}$ on the generalized $n$-qubit Bloch sphere. Dividing the
tally $p^{2^{n}-1}(p^{2^{n}}-1)$ of unit norm states by the $p+1$
elements of each phase-equivalent discrete circle, we find 
\[
\frac{p^{2^{n}-1}(p^{2^{n}}-1)}{p+1}=p^{2^{n}-1}(p-1)\,\prod_{k=1}^{n-1}(p^{2^{k}}+1)
\]
as the total count of unique irreducible states in a discrete $n$-qubit
configuration (see \ref{appI.sec}). The resulting object is precisely
the discrete version of $\CP{D-1}$, which we might call a \textit{discrete
complex projective space\/} or $\DCP{D-1}$, where $D=2^{n}$ as
usual.

@@

Finally, we construct the discrete version of the quadratic unit-length
form that automatically annihilates the distinction among states differing
only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Vec{X}=\{|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\}\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Vec{X}\cdot\Vec{X}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
Computing all the elements of the discrete $\Sphere{2\times2^{n}-1}$
structure that map to the \textit{same point\/} in $\Vec{X}$, we
find a $(p+1)$-fold redundancy, exactly as for the one-qubit case;
each set of $(p+1)$ redundant points is, geometrically speaking,
the \textit{discrete Hopf fibration circle\/} (see, e.g., Arnold
\cite{Arnold}) living above each \textit{irreducible\/} point of
the $n$-qubit state description.

In summary, when the $n$-qubit continuous Hopf fibration process
is repeated for the discrete analog of $n$-qubit states, we find
that that analog of a phase ``circle'' is a discrete set $p+1$
equivalent points that map to the same point $\Vec{X}$ on the generalized
$n$-qubit Bloch sphere, and therefore there are a total of $p^{2^{n}-1}(p^{2^{n}}-1)/(p+1)$
unique irreducible states in a discrete $n$-qubit configuration.
The resulting object is precisely the discrete version of $\CP{D-1}$,
which we might call a \textit{discrete complex projective space\/}
or $\DCP{D-1}$. }

>>>>

Thus the geometry of a single qubit reduces to transformations among
points on $\Sphere{2}$, which can be parametrized in an infinite
one-parameter family of transformations, one of which is the geodesic
or minimal-length transformation. Explicitly, given two one-qubit
states denoted by points $\Hat{a}$ and $\Hat{b}$ on $\Sphere{2}$,
the shortest rotation carrying $\Hat{a}$ to $\Hat{b}$ is the SLERP
(spherical linear interpolation) \cite{Shoemake:1985:ARQ:325334.325242,wiki:Slerp}
\begin{equation}
S\left(\Hat{a},\Hat{b},t\right)=\Hat{a}\,\frac{\sin((1-t)\omega)}{\sin\omega}\,+\,\Hat{b}\,\frac{\sin(t\omega)}{\sin\omega}\,,\label{slerp.eq}
\end{equation}
where $\Hat{a}\cdot\Hat{b}=\cos\omega$. Figure \ref{blochSphere.fig}(b)
illustrates the path traced by a SLERP between two irreducible one-qubit
states on the Bloch sphere. Because states in CQC are defined by infinite
precision real numbers, it is not possible, even in principle, to
make an exact state transition as implied by Fig.~\ref{blochSphere.fig}(b).
In practice, one has to be content with approximate, typically exponentially
expensive, transitions from state to state.

@@

\subsection{Explicit generalization of the Hopf fibration construction}

For a two-dimensional system, we could easily solve the problem of
reducing the full unit-norm space to its irreducible components $\Hat{a}=\left(X,Y,Z\right)$
characterizing the Bloch sphere. We have just argued that essentially
the same process is possible for $D$-dimensional system: in the abstract
argument, we simply identify the family of coefficients $\left\{ \alpha_{i}\right\} $
as being the same if they differ only by an overall phase $\rme^{\rmi\theta}$.
However, in practice this is not a construction that is easy to realize
in a practical computation. We now outline an explicit algorithm for
accomplishing the reduction to the irreducible $D$-dimensional state
space $\CP{D-1}$; this construction will turn out to be useful for
the validation of our discrete results to follow below.

Given a normalized pure state $\ket{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}\ket{i}$,
a natural quantity characterizing an $D$-dimensional system is its
\textit{density matrix}, $\rho=\op{\Psi}{\Psi}$, or 
\begin{equation}
\rho=\begin{bmatrix}|\alpha_{0}|^{2} & \alpha_{0}\alpha_{1}^{*} & \cdots & \alpha_{0}\alpha_{D-1}^{*}\\
\alpha_{1}\alpha_{0}^{*} & |\alpha_{1}|^{2} & \cdots & \alpha_{1}\alpha_{D-1}^{*}\\
\vdots & \vdots & \ddots & \vdots\\
\alpha_{D-1}\alpha_{0}^{*} & \cdots & \alpha_{D-1}\alpha_{D-2}^{*} & |\alpha_{D-1}|^{2}
\end{bmatrix}\,.\label{nqDensity.eq}
\end{equation}
We can now use the complex generalization of the classical Veronese
coordinate system for projective geometry to remove the overall phase
ambiguity $\rme^{\rmi\theta}$ from the $D$-dimensional states. If
we take a particular weighting of the elements of the density matrix~$\rho$,
we can construct a \textit{unit vector} of real dimension $D^{2}$
with the form: 
\begin{equation}
\Hat{a}=\left(|\alpha_{i}|^{2},\ldots,\sqrt{2}\,\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\,\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\right)\,,\label{nqCoords.eq}
\end{equation}
where 
\begin{equation}
\Hat{a}\cdot\Hat{a}=\sum_{i=0}^{D-1}\left(\left|\alpha_{i}\right|^{2}\right)^{2}+\sum_{i=0}^{D-1}\sum_{\substack{j=0\\
j\ne i
}
}^{D-1}\left(\mathrm{Re}\ \alpha_{i}\alpha_{j}^{*}\right)^{2}+\left(\mathrm{Im}\ \alpha_{i}\alpha_{j}^{*}\right)^{2}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)\left(\sum_{j=0}^{D-1}\left|\alpha_{j}\right|^{2}\right)=1\,.
\end{equation}
This construction gives an explicit embedding of the $\left(D-1\right)$-dimensional
complex, or $\left(2D-2\right)$-dimensional real, object in a real
space of dimension $D^{2}$. However, this is somewhat subtle because
the vector is of unit length, so technically the embedding space is
a sphere of dimension $D^{2}-1$ embedded in $\R^{D^{2}}$. For example,
the two-dimensional irreducible states could be represented in a four-dimensional
embedding, but the magnitude of every coordinate would be one; furthermore,
the object embedded in the resulting $\Sphere{3}$ is indeed $\Sphere{2}$
because we can fix one complex coordinate to be unity, and let one
vary, giving a total of two irreducible dimensions. In fact one must
choose \textit{two} coordinate patches, one covering one pole of $\Sphere{2}$
with coordinates $\alpha_{0}=1+\rmi0$ and $\alpha_{1}=x_{1}+\rmi y_{1}$,
and the other patch covering the other pole of $\Sphere{2}$ with
coordinates $\alpha_{0}=x_{0}+\rmi y_{0}$ and $\alpha_{1}=1+\rmi0$.

We finally see that the irreducible $D$-dimensional state space $\CP{D-1}$
is described by $D$ projectively equivalent coordinates, one of which
can always be scaled out to leave $(D-1)$ actual (complex) degrees
of freedom. We must choose, in turn, $D$ different local sets of
complex variables defined by taking the value $\alpha_{k}=1$, with
$k=0,\ldots,D-1$, and allowing the remaining $D-1$ complex (or $2D-2$
real) variables to run free. No single set of coordinates will work,
since the submanifold including $\alpha_{k}=0$ is undefined and another
coordinate system must be chosen to cover that coordinate patch. This
is a standard feature of the topology of non-trivial manifolds such
as $\CP{D-1}$ (see any textbook on geometry~\cite{Berger_1988}).

@@

This process of describing the discrete $n$-qubit irreducible states
can again be understood geometrically by following the discrete analog
of the Hopf fibration. First, we construct the discrete version of
the quadratic unit-length form that automatically annihilates the
distinction among states differing only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=(|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
{}From \ref{appH.sec}, we know that $p+1$ elements of this discrete
$\Sphere{2\times2^{n}-1}$ structure map to the \textit{same point\/}
in $\Hat{a}$. Each set of $(p+1)$ redundant points is, geometrically
speaking, the \textit{discrete Hopf fibration circle\/} living above
each \textit{irreducible\/} point of the $n$-qubit state description.
These $p+1$ points are interpretable as the $p$ finite points plus
the single point at infinity of the projective discrete line (see,
e.g., \cite{Arnold}).

@@

Finally, we repeat the last step of the $n$-qubit continuous Hopf
fibration process for discrete $n$-qubit states, eliminating the
discrete set of $p+1$ equivalent points that map to the same point
$\Hat{a}$ on the generalized $n$-qubit Bloch sphere. Dividing the
tally $p^{2^{n}-1}(p^{2^{n}}-1)$ of unit norm states by the $p+1$
elements of each phase-equivalent discrete circle, we find 
\[
\frac{p^{2^{n}-1}(p^{2^{n}}-1)}{p+1}=p^{2^{n}-1}(p-1)\,\prod_{k=1}^{n-1}(p^{2^{k}}+1)
\]
as the total count of unique irreducible states in a discrete $n$-qubit
configuration (see \ref{appI.sec}). The resulting object is precisely
the discrete version of $\CP{D-1}$, which we might call a \textit{discrete
complex projective space\/} or $\DCP{D-1}$, where $D=2^{n}$ as
usual.

@@

Finally, we construct the discrete version of the quadratic unit-length
form that automatically annihilates the distinction among states differing
only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=\{|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\}\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
Computing all the elements of the discrete $\Sphere{2\times2^{n}-1}$
structure that map to the \textit{same point\/} in $\Hat{a}$, we
find a $(p+1)$-fold redundancy, exactly as for the one-qubit case;
each set of $(p+1)$ redundant points is, geometrically speaking,
the \textit{discrete Hopf fibration circle\/} (see, e.g., Arnold
\cite{Arnold}) living above each \textit{irreducible\/} point of
the $n$-qubit state description.

In summary, when the $n$-qubit continuous Hopf fibration process
is repeated for the discrete analog of $n$-qubit states, we find
that that analog of a phase ``circle'' is a discrete set $p+1$
equivalent points that map to the same point $\Hat{a}$ on the generalized
$n$-qubit Bloch sphere, and therefore there are a total of $p^{2^{n}-1}(p^{2^{n}}-1)/(p+1)$
unique irreducible states in a discrete $n$-qubit configuration.
The resulting object is precisely the discrete version of $\CP{D-1}$,
which we might call a \textit{discrete complex projective space\/}
or $\DCP{D-1}$. }
